Optimal management of a stochastically varying population when policy adjustment is costly

### 2014-05-23 manuscripts/manuscript.Rmd
avoid printing unnecessary outputs

----
title: Optimal management of a stochastically varying population when policy adjustment is costly
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Michael Bode
    affiliation: melbourne
  - name: Jacob LaRiviera
    affiliation: UTK
  - name: James Sanchirico
    affiliation: UCD
  - name: Alan Hastings
    affiliation: UCD
  - name: Paul Armsworth 
    affiliation: UTK

address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
  - code: UCD
    address: Department of Environmental Science and Policy, University of California, Davis
  - code: melbourne
    address: University of Melbourne, Australia 
  - code: UTK
    address: University of Tennessee, Knoxville

abstract: |
          Ecological systems are dynamic and policies to manage them need to
          respond to that variation. However, policy adjustments will sometimes
          be costly, which means that fine-tuning a policy to track variability in
          the environment very tightly will only sometimes be worthwhile. We use a
          classic fisheries management question – how to manage a stochastically
          varying population using annually varying quotas in order to maximize
          profit  - to examine how costs of policy adjustment change optimal
          management recommendations. Costs of policy adjustment (here changes in
          fishing quotas through time) could take different forms. For example,
          these costs may respond to the size of the change being implemented or
          there could be a fixed cost any time a quota change is made. We show
          how different forms of policy costs have contrasting implications for
          optimal policies. While some types of cost act to smooth out variation in
          quotas and stock sizes, as one might have expected, others can actually
          increase variation in stock sizes and quotas through time. We also show
          that the potential economic impact on the fishery of managers assuming
          policy adjustment costs are present when in fact they are absent is much
          smaller than the impact of managers assuming such costs are absent if
          they are present.

layout: preprint      # review, 11pt
---

>>>>

---
title: Optimal management of a stochastically varying population when policy adjustment is costly
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Michael Bode
    affiliation: melbourne
  - name: Jacob LaRiviera
    affiliation: UTK
  - name: James Sanchirico
    affiliation: UCD
  - name: Alan Hastings
    affiliation: UCD
  - name: Paul Armsworth 
    affiliation: UTK
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
  - code: UCD
    address: Department of Environmental Science and Policy, University of California, Davis
  - code: melbourne
    address: University of Melbourne, Australia 
  - code: UTK
    address: University of Tennessee, Knoxville
abstract: |
          Ecological systems are dynamic and policies to manage them need to
          respond to that variation. However, policy adjustments will sometimes
          be costly, which means that fine-tuning a policy to track variability in
          the environment very tightly will only sometimes be worthwhile. We use a
          classic fisheries management question – how to manage a stochastically
          varying population using annually varying quotas in order to maximize
          profit  - to examine how costs of policy adjustment change optimal
          management recommendations. Costs of policy adjustment (here changes in
          fishing quotas through time) could take different forms. For example,
          these costs may respond to the size of the change being implemented or
          there could be a fixed cost any time a quota change is made. We show
          how different forms of policy costs have contrasting implications for
          optimal policies. While some types of cost act to smooth out variation in
          quotas and stock sizes, as one might have expected, others can actually
          increase variation in stock sizes and quotas through time. We also show
          that the potential economic impact on the fishery of managers assuming
          policy adjustment costs are present when in fact they are absent is much
          smaller than the impact of managers assuming such costs are absent if
          they are present.
layout: preprint      # review, 11pt
---

<!-- CODE -->

```{r libraries, output="hide", include=FALSE, cache=FALSE}
rm(list=ls())
library("pdgControl")
library("reshape2")
library("ggplot2")
library("data.table")
library("pander")
theme_set(theme_bw())
```

``` {r profit_model}
price = 10
c0 = 30
profit <- profit_harvest(price = price, c0 = c0, c1 = 0)
```

```{r c2_grid}
c2 <- exp(seq(0, log(41), length.out = 40))-1
c2 <- seq(0, 40, length.out=100)
```

```{r reduction}
reduction <- 0.25 
```

```{r setup}
seed <- 123                 # Random seed (replicable results)
delta <- 0.05               # economic discounting rate
OptTime <- 20               # stopping time
gridsize <- 50              # grid size for fish stock and harvest rate (discretized population)
sigma_g <- 0.2              # Noise in population growth
reward <- 0                 # bonus for satisfying the boundary condition
z_g <- function() rlnorm(1,  0, sigma_g) # mean 1
z_m <- function() 1         # No measurement noise, 
z_i <- function() 1         # No implemenation noise
f <- BevHolt                # Select the state equation
pars <- c(1.5, 0.05)        # parameters for the state equation
K <- (pars[1] - 1)/pars[2]  # Carrying capacity (for reference 
xT <- 0                     # boundary conditions
x0 <- K
x_grid <- seq(0.01, 1.2 * K, length = gridsize)  
h_grid <- seq(0.01, 0.8 * K, length = gridsize)  
````

```{r reed, dependson=c("setup", "profit_model")}
SDP_Mat <- determine_SDP_matrix(f, pars, x_grid, h_grid, sigma_g )
opt <- find_dp_optim(SDP_Mat, x_grid, h_grid, OptTime, xT, 
                     profit, delta, reward=reward)
````

```{r fees}
L1 <- function(c2) function(h, h_prev)  c2 * abs(h - h_prev) 
fixed <-  function(c2) function(h, h_prev) c2 * as.numeric( !(h == h_prev) )
L2 <- function(c2) function(h, h_prev)  c2 * (h - h_prev) ^ 2
none <- function(h, h_prev)  0
penaltyfns <- list(L2=L2, L1=L1, fixed=fixed)
````

```{r parallel, include=FALSE}
require(snowfall)
sfInit(cpu=4, parallel=T)
sfLibrary(pdgControl)
sfExportAll()
````

``` {r bigloop, dependson=c("setup", "reed", "fees", "profit_model", "c2_grid")}
policies <- lapply(penaltyfns, function(penalty){
  sfLapply(c2, function(c2){
      policy <- optim_policy(SDP_Mat, x_grid, h_grid, OptTime, xT, 
                   profit, delta, reward, penalty = penalty(c2))
      }
  )
})
````

```{r, dependson="quadcosts"}
i <- which(x_grid > K)[1]
fees <- 
lapply(policies, function(penalty) 
  sapply(penalty, function(c2_run)
    max(c2_run$V[i,]) # Would be penalty_free_V originally   ## this isn't correct for asym cases
  )
)
```

``` {r npv-plot, dependson="quadcosts"}
npv0 <- max(fees$L1) # all have same max, at c2=0 
fees <- data.frame(c2=c2,fees)
fees <- melt(fees, id="c2")

fees <- subset(fees, variable %in% c("L1", "L2", "fixed"))
ggplot(fees, aes(c2, value, col=variable)) + geom_point() + geom_line()
````

```{r apples_plot, dependson="quadcosts"}
ggplot(fees, aes(c2, (npv0-value)/npv0, col=variable)) + geom_point() + geom_line()
````

```{r apples, dependson=c("quadcosts", "reduction")}
closest <- function(x, v){
  which.min(abs(v-x))
}
dt_npv <- data.table(fees)
index <- dt_npv[,closest(reduction, (npv0-value)/npv0), by=variable]
apples_index <- index$V1
names(apples_index) = index$variable
apples <- c2[index$V1]
````

```{r reductions}
reduction_table <- sapply(c("0.10" = 0.10, "0.20" = 0.20, "0.30" = 0.30), function(reduction){
index <- dt_npv[,closest(reduction, (npv0-value)/npv0), by=variable]
apples <- c2[index$V1]
names(apples) = index$variable
apples })
write.csv(reduction_table, "reduction_table.csv")
````

```{r print_npv, dependson="apples"}
setkey(dt_npv, variable)
values <- apply(index, 1, function(x) dt_npv[x[1], ][as.integer(x[2]), ]$value)
percent.error <- (values - ((1-reduction)*npv0)) / ((1-reduction)*npv0)* 100
print_npv <- data.frame(model=index$variable, "value realized"=values, 
                        "percent of npv0" = 100*values/npv0, "percent error"=percent.error)

```

```{r npv_table, dependson="print_npv", results='asis'} 
pandoc.table(print_npv)
```

```{r policynames}
L2_policy <- policies$L2[[apples_index["L2"]]]$D
L1_policy <- policies$L1[[apples_index["L1"]]]$D
fixed_policy <- policies$fixed[[apples_index["fixed"]]]$D
```

```{r simulate_policy, dependson=c("policynames", "apples")}
reps <- 1:100
names(reps) = paste("rep", 1:length(reps), sep="_") # treat as a factor
seeds <- 1:100
sims <- list(
  L1 = lapply(reps, function(x) simulate_optim(f, pars, x_grid, h_grid, x0, 
                      L1_policy, z_g, z_m, z_i, 
                      opt$D, profit=profit, penalty=L1(apples["L1"]), seed=seeds[x])), 
  L2 = lapply(reps, function(x) simulate_optim(f, pars, x_grid, h_grid, x0, 
                      L2_policy, z_g, z_m, z_i, 
                      opt$D, profit=profit, penalty=L2(apples["L2"]), seed=seeds[x])),
  fixed = lapply(reps, function(x) simulate_optim(f, pars, x_grid, h_grid, x0, 
                         fixed_policy, z_g, z_m, z_i, 
                         opt$D, profit=profit, penalty=fixed(apples["fixed"]), seed=seeds[x]))

)
```

```{r tidy, dependson="simulate_policy"}
#Make data tidy (melt), fast (data.tables), and nicely labeled.
dat <- melt(sims, id=names(sims[[1]][[1]]))  
dt <- data.table(dat)
setnames(dt, "L2", "replicate") # names are nice
setnames(dt, "L1", "penalty_fn") # names are nice
```

```{r profit_calcs, dependson="tidy"}
# Profit when accounting for penalty when present
optimal_cost <- dt[, sum(profit_fishing - policy_cost), by=penalty_fn ]

# Profit when ignoring penalty when present
ignore_when_present <- dt[, sum(profit_fishing_alt - policy_cost_alt), by=penalty_fn]

# Profit when assuming penalty when it is absent
assume_when_absent <- dt[, sum(profit_fishing), by=penalty_fn]

# Profit when ignoring penalty when it is absent
optimal_free <- dt[, sum(profit_fishing_alt), by=penalty_fn]

# Normalize by the optimal 
ignore_fraction <- ignore_when_present$V1/optimal_free$V1 # common normalization
assume_fraction <- assume_when_absent$V1/optimal_free$V1
assume_when_absent <- cbind(assume_when_absent, assume_fraction = assume_fraction, normalize_optimal_free=optimal_free$V1, normalize_optimal_cost = optimal_cost$V1)
ignore_when_present <- cbind(ignore_when_present, ignore_fraction = ignore_fraction)

# Name and merge columns
setnames(ignore_when_present, "V1", "ignore_cost")
setnames(assume_when_absent, "V1", "assume_cost")
error_costs <- merge(ignore_when_present, assume_when_absent, "penalty_fn")

# print_npv                             # theoretically acheivable profits under these costs
# optimal_cost$V1/optimal_free$V1       # actually realized

error_costs <- cbind(error_costs, sigma_g = sigma_g, reduction = reduction)

```

```{r}
v <- dt[,var(harvest), by="penalty_fn"]
var <- v$V1
names(var) <- v$penalty_fn
acor <- dt[,acf(harvest, plot=F)$acf[2], by="penalty_fn"]$V1
names(acor) <- names(var)
out <- rbind(var=var, a=acor)
```

```{r p1, dependson="tidy", fig.width=12}
p1 <- ggplot(subset(dt, replicate %in% names(reps)[1:5])) +
  geom_line(aes(time, alternate), col="grey20", lwd=1) +
  geom_line(aes(time, fishstock), col=rgb(0,0,1,.8)) + 
  facet_grid(replicate~penalty_fn) + 
  labs(x="time", y="stock size", title = "Stock Dynamics")
p1
```

```{r Figure3, dependson="tidy", fig.width=12}
p2 <- ggplot(subset(dt, replicate %in% names(reps)[1:5])) +
  geom_line(aes(time, harvest_alt), col="grey20", lwd=1)  +
  geom_line(aes(time, harvest), col=rgb(0,0,1,.8)) + 
  facet_grid(replicate~penalty_fn) + 
  labs(x="time", y="havest intensity (fish taken)", title = "Harvest Policy Dynamics")
p2
````

```{r}
frac_lost <- seq(0,1, length=20)
```

```{r helper_fn_1}
fig4 <- function(fraction_lost){
closest <- function(x, v){
  which.min(abs(v-x))
}
dt_npv <- data.table(fees)
index <- dt_npv[,closest(fraction_lost, (npv0-value)/npv0), by=variable]
apples_index <- index$V1
names(apples_index) = index$variable
apples <- c2[index$V1]
names(apples) = index$variable

L2_policy <- policies$L2[[apples_index["L2"]]]$D
L1_policy <- policies$L1[[apples_index["L1"]]]$D
fixed_policy <- policies$fixed[[apples_index["fixed"]]]$D
free_increase_policy <- policies$free_increase[[apples_index["free_increase"]]]$D
free_decrease_policy <- policies$free_decrease[[apples_index["free_decrease"]]]$D
quad_policy <- policies$quad[[apples_index["quad"]]]$D

quad_profit <- profit_harvest(price = price, c0 = c0, c1 = apples["quad"]) 
sims <- lapply(1:50, function(reps) list(
  L1 = simulate_optim(f, pars, x_grid, h_grid, x0, 
                      L1_policy, z_g, z_m, z_i, 
                      opt$D, profit=profit, penalty=L1(apples["L1"])), 
  L2 = simulate_optim(f, pars, x_grid, h_grid, x0, 
                      L2_policy, z_g, z_m, z_i, 
                      opt$D, profit=profit, penalty=L2(apples["L2"])),
  fixed = simulate_optim(f, pars, x_grid, h_grid, x0, 
                         fixed_policy, z_g, z_m, z_i, 
                         opt$D, profit=profit, penalty=fixed(apples["fixed"]))
#  increase = simulate_optim(f, pars, x_grid, h_grid, x0, 
#                            free_increase_policy, z_g, z_m, z_i, 
#                            opt$D, profit=profit, penalty= free_increase(apples["increase"])),
#  decrease = simulate_optim(f, pars, x_grid, h_grid, x0, 
#                            free_decrease_policy, z_g, z_m, z_i, 
#                            opt$D, profit=profit, penalty= free_decrease(apples["decrease"])),
#  quad = simulate_optim(f, pars, x_grid, h_grid, x0, 
#                            quad_policy, z_g, z_m, z_i, 
#                            opt$D, profit=quad_profit, penalty= none)
  ))

  #Make data tidy (melt), fast (data.tables), and nicely labeled.
  dat <- melt(sims, id=names(sims[[1]][[1]]))  
  dt <- data.table(dat)
  setnames(dt, "L1", "reps") # names are nice
  setnames(dt, "L2", "penalty_fn") # names are nice

  dt
}
```

```{r helper_fn_2} 
# Helper functions to extract the summary stats in different variables
stats_harvest <- function(dt){
v <- dt[,var(harvest), by=c("penalty_fn", "reps")]
acor <- dt[,acf(harvest, plot=F)$acf[2], by=c("penalty_fn", "reps")]
df <- cbind(v, acor$V1)
setnames(df, c("V1", "V2"), c("var", "acor")) # names are nice
df
}

stats_fishstock <- function(dt){
v <- dt[,var(fishstock), by=c("penalty_fn", "reps")]
acor <- dt[,acf(fishstock, plot=F)$acf[2], by=c("penalty_fn", "reps")]
df <- cbind(v, acor$V1)
setnames(df, c("V1", "V2"), c("var", "acor")) # names are nice
df
}
```

```{r helper_fn_3}
#' a simple function for reorganizing the data over the different "faction-lost" levels
get_trends <- function(tmp2){
  out <- melt(tmp2, id=c("reps", "var", "acor"))
  colnames(out) = c("reps", "var", "acor", "nothing", "penalty", "index")
  out <- cbind(out[c(1,2,3,5)], fraction = frac_lost[out$index])
  out <- data.table(out)
  Ev = out[,mean(var),by=c('penalty', 'fraction')]
  SDv = out[,sd(var),by=c('penalty', 'fraction')]
  Ea = out[,mean(acor),by=c('penalty', 'fraction')]
  SDa = out[,sd(acor),by=c('penalty', 'fraction')]
  harvest_trends <- data.table(penalty = Ev$penalty, 
                               fraction = Ev$fraction, 
                               Ev = Ev$V1, Ea = Ea$V1, 
                               SDv=SDv$V1, SDa = SDa$V1)
}
```

```{r Figure4_harvest}
sims_at_each_apple <- lapply(frac_lost, fig4)
harvest_stats <- lapply(sims_at_each_apple, stats_harvest)
harvest_trends <- get_trends(harvest_stats)
```

```{r Figure4}
Fig4a <- ggplot(harvest_trends, 
                aes(fraction, Ev, ymin=Ev-SDv, ymax=Ev+SDv, col=penalty)) + 
  geom_ribbon(aes(fill=penalty, col=NA), lwd=0, alpha=.05) + 
  geom_line() + 
  xlab("Fraction of NPV lost to costs") +
  scale_x_continuous(limits=c(0, 0.4))

Fig4b <- ggplot(harvest_trends, 
                aes(fraction, Ea, ymin=Ea-SDa, ymax=Ea+SDa, col=penalty)) + 
  geom_ribbon(aes(fill=penalty, col=NA), lwd=0, alpha=.05) + 
  geom_line() + xlab("Fraction of NPV lost to costs")+
  scale_x_continuous(limits=c(0, 0.4)) 
Fig4a
Fig4b
````

```{r Figure4S}
fishstock_stats <- lapply(sims_at_each_apple, stats_fishstock)
fishstock_trends <- get_trends(fishstock_stats)

FigS4a <- ggplot(fishstock_trends, 
                aes(fraction, Ev, ymin=Ev-SDv, ymax=Ev+SDv, col=penalty)) + 
  geom_ribbon(aes(fill=penalty, col=NA), lwd=0, alpha=.05) + 
  geom_line() + xlab("Fraction of NPV lost to costs")+
  scale_x_continuous(limits=c(0, 0.4))

FigS4b <- ggplot(fishstock_trends, 
                aes(fraction, Ea, ymin=Ea-SDa, ymax=Ea+SDa, col=penalty)) + 
  geom_ribbon(aes(fill=penalty, col=NA), lwd=0, alpha=.05) + 
  geom_line() + xlab("Fraction of NPV lost to costs") +
  scale_x_continuous(limits=c(0, 0.4)) 
FigS4a
FigS4b
```

<!-- ENDCODE -->
